

\chapter*{Preface}\normalsize
  \addcontentsline{toc}{chapter}{Preface}

The primary objective of this research effort is to develop desicion making framework in the design and operation of Power Systems.  The methodology involved will be derived from the Industrial Engineering discipline of Operations Research.  Thanks to all those who came before to develop the tools and methods I could never dream of.

Thanks to UW-Madison for providing the opportunity to learn and grow.  In particular, to the many Professors I have had the privelidge to work with and who have tought me the tools of Operations Research.   Thanks to Jeffrey Linderoth for advising me along the way and helping me find a path (as well as all the knowledge about things like Stochastic Programming and slowly converting me to the glory of linux workflows); Stephen Robinson for somehow always being able to explain theoretical concepts so they make sense and combining topics such as economics, probability, and optimization in applied settings seemlessly; Jim Luedtke for helping me find a place in the department and wisdom in integer programming; Oguzhan Alagoz for knowledge in Markov Decision Processes; and Vicki Bier for topics relating to risk analysis.  Thanks to Bernard Lesieutre for helping me understand power systems and Gregory Nemet for issues around energy policy.

Thanks to Jeff Twidwell, Kevin Rockwell, my Parents, and others for helping me edit this document to make it coherent and powerful.

Thanks to my Mom (Sue), Dad (Tim), and Sister (Steph) for always giving me a supportive environment to develop as a person.  Thanks to all my extended family and friends for helping me enjoy life.

